Composable Generalized Consensus

### 2015-05-20 paper/abstract3.tex
Submitted version

The key novelty of our framework is its practical applicability. 
Our framework is based on Generalized Paxos, therefore providing buit-in support
for algorithms optimizing the execution of commuting commands.
Moreover, we propose two specifications 
capturing the essence of modular SMR algorithms based on quorum systems.
We demonstrate that concrete
algorithms can easily be obtained by refinement from our abstract specifications, as
evidenced by TLA+ specifications ready to be checked by the TLC model checker.

}

>>>>

The key novelty of our framework is its practical applicability: we show that
a large class of SMR algorithms can be expressed in our framework by proposing
two specifications capturing the essence of modular SMR algorithms based
on Generalized Consensus and using quorum systems for synchronization. We
demonstrate that concrete algorithms can easily be obtained by refinement from
our abstract specifications, as evidenced by TLA+ specifications ready to be
checked by the TLC model checker.
}

### 2015-05-20 paper/intro.tex
Submitted version

The key novelty of our framework lies in its practical applicability.
Most SMR algorithms are variations on a basic structure consisting of one of two kinds of 
quorum systems \cite{GuerraouiVukolic10RefinedQuorumSystems}. 
The first kind, that we call resilient quorums, are used in leader-based algorithms, while the second
kind of quorums, that we call fast quorums, are used in leader-less algorithms and can order commuting commands without synchronization.
We provide two specifications, at an intermediate level of abstraction, that capture the essence of composing algorithms using these two kinds of quorums.
Our two specifications allow to easily check that new modules based on resilient or fast quorums are correct, covering a large spectrum of practical algorithms.

Allowing the execution of commuting commands to proceed without
synchronization is an important practical optimization, as commands may
often commute in practice, for example when operating on disjoint parts
of a data structure. To enable the synchronization-free execution of
commuting commands, we have based our framework on Generalized Consensus
\cite{Lamport05GeneralizeConsensus}. Traditional SMR implementations use a
separate instance of a consensus algorithm to agree on the command to execute
at each index of the command sequence on which the replicas must agree. Using
separate consensus instances makes it hard to optimize the execution of
commuting command because commutativity is not a property of a single command.
Instead of restricting agreement to a single value, the Generalized Consensus
problem requires that set of replica servers agree on the state of a growing
data structure called a command-structure set. A command-structure set is
a data structure with an append operation such that the order in which two
commuting commands are appended does not change the resulting state. Thanks to
this property, two replicas executing commuting commands in a different order
will not end up in a different state and can continue executing without further
synchronization. In message-passing systems, this property can be leveraged to
reduce the latency of processing one command to two message delays instead of
three. As we will show, it also enables the processing of read-only commands
in two message delays, although read-only commands do not commute with write
commands.

In practice, a SMR algorithm must support fast reconfiguration to replace faulty
components, without which the long-term stability of a system is inevitably
compromised. Our framework abstracts over how the reconfiguration decisions
are made, providing an interface allowing to plug an external reconfiguration
master implementing the reconfiguration policy of the system. The master decides
which GC module to run next and which replica servers will participate, removing
failed replicas and adding new ones. As we will explain, this enables tolerating
$k$ failures with $k+1$ replicas, an important improvement over the traditional
$2k+1$ needed replicas.

To demonstrate the practical applicability of our framework we provide
composable versions of the generalized variations of popular consensus
algorithms: classic Paxos, Fast Paxos, Chain Replication, and multi-coordinated
Paxos. We have checked the correctness of these algorithms with the TLC
model-checker and we provide all the specifications needed to replicate these
experiments. Our framework is also fully formalized and its composition theorem
proved in Isabelle/HOL, a foundational interactive theorem prover, thereby
certifying the correctness of the composition theorem at the deepest level.

At the heart of our framework lies an abstraction, called Composable Generalized
Consensus, of the familiar notion of round (also called epoch, ballot, or
view in the literature) used in most SMR algorithms. As an example, let us
consider Fast Paxos, one of the first SMR algorithm which can change strategy
dynamically. Fast Paxos is a round-based SMR algorithm that has two types of
rounds: fast rounds and normal rounds. Fast rounds process requests faster
than normal rounds in contention-free executions but fall back to a slower
scheme under contention. Normal rounds are the same as in the classic Paxos
and are slower than fast rounds in contention-free executions. Any process can
initiate a round change at any point during execution: this allows switching
between fast and normal rounds as well as changing the leader from one normal
round to the next, similarly to classic Paxos. Each kind of round maintains
different invariants relating the state of each process and the values that
might have been chosen in the previous rounds, thus a given round must know
which past rounds are fast or normal rounds in order to take over the execution
safely. When adding more kinds of rounds such an approach will quickly become
unmanageable, as each type of rounds must have one case for each possible other
kind of round, making $O(n^2)$ cases to consider in total.

Our notion of Composable Generalized Consensus defines a common interface and a
common correctness property for a round, abstracting over the particular round
implementations and making all types of rounds compatible with each other.
Moreover, our interface for rounds enables a crucial optimization seen in
practical systems: introducing an external reconfiguration master which manages
the addition of new replicas and the removal of failed ones. Interestingly, as
explained in \cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication},
in a real system it is often possible to implement a reliable reconfiguration
master which allows to tolerate $f$ faults with $f+1$ replicas instead of the
traditional $2f+1$ replicas. The reconfiguration master can also make the choice
of the new type of round to execute, providing the system's reconfiguration
policy.

>>>>

The key novelty of our framework lies in its practical applicability:
\begin{itemize}
  \item Our framework is based on Generalized Consensus, a generalization of the consensus problem 
    that allows optimizing the execution of commuting commands.
  \item Many SMR algorithms are variations on a basic structure consisting of one of two kinds of 
    quorum systems \cite{GuerraouiVukolic10RefinedQuorumSystems}. 
    The first kind, that we call resilient quorums, are used in leader-based algorithms, while the second
    kind of quorums, that we call fast quorums, are used in leader-less algorithms and can order commuting commands without synchronization.
    We provide two specifications, at an intermediate level of abstraction, that capture the essence of composing algorithms using these two kinds of quorums.
    Our two specifications allow to easily check that new modules based on resilient or fast quorums are correct, covering a large spectrum of practical algorithms. 
\end{itemize}

Allowing the execution of commuting commands to proceed without
synchronization is an important practical optimization, as commands may
often commute in practice, for example when operating on disjoint parts
of a data structure. To enable the synchronization-free execution of
commuting commands, we have based our framework on Generalized Consensus
\cite{Lamport05GeneralizeConsensus}. Traditional SMR implementations use a
separate instance of a consensus algorithm to agree on the command to execute
at each index of the command sequence on which the replicas must agree. Using
separate consensus instances makes it hard to optimize the execution of
commuting command because commutativity is not a property of a single command.
Instead of restricting agreement to a single value, the Generalized Consensus
problem requires that set of replica servers agree on the state of a growing
data structure called a command-structure set. A command-structure set is
a data structure with an append operation such that the order in which two
commuting commands are appended does not change the resulting state. Thanks to
this property, two replicas executing commuting commands in a different order
will not end up in a different state and can continue executing without further
synchronization. In message-passing systems, this property can be leveraged to
reduce the latency of processing one command to two message delays instead of
three. 
%As we will see, fast execution of commutative requests also enables the processing of read-only commands
%in two message delays, although read-only commands do not commute with write
%commands.

In practice, a SMR algorithm must support fast reconfiguration to replace
faulty components, without which the long-term stability of a system is
inevitably compromised. Our framework abstracts over how the reconfiguration
decisions are made, providing an interface allowing to plug in an external
reconfiguration master implementing the reconfiguration policy of the
system. The master decides which GC module to run next and which replica
servers will participate, removing failed replicas and adding new ones. The
reconfiguration master enables tolerating $k$ failures with $k+1$ replicas
\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}, an important
improvement over the traditional $2k+1$ needed replicas.

To demonstrate the practical applicability of our framework we provide
composable versions of the generalized variations of popular consensus
algorithms: Classic Paxos, Fast Paxos, Chain Replication, and multi-coordinated
Paxos. Thanks to our specifications of algorithms based on quorum-systems, 
we have checked the correctness of our version of Fast Paxos and Chain Replication with the TLC
model-checker and we provide all the specifications needed to replicate these
experiments. Our framework is also fully formalized and its composition theorem
proved in Isabelle/HOL, a foundational interactive theorem prover, thereby
certifying the correctness of the composition theorem at the deepest level.
The TLA+ and Isabelle/HOL development can be found at \url{http://losa.fr/cgc}.

At the heart of our framework lies an abstraction, called Composable Generalized
Consensus (CGC), of the familiar notion of round (also called epoch, ballot, or
view in the literature) used in most SMR algorithms. As an example, let us
consider Fast Paxos, one of the first SMR algorithm which can change strategy
dynamically. Fast Paxos is a round-based SMR algorithm that has two types of
rounds: fast rounds and normal rounds. Fast rounds process requests faster
than normal rounds in contention-free executions but fall back to a slower
scheme under contention. Normal rounds are the same as in the classic Paxos
and are slower than fast rounds in contention-free executions. Any process can
initiate a round change at any point during execution: this allows switching
between fast and normal rounds as well as changing the leader from one normal
round to the next, similarly to classic Paxos. Each kind of round maintains
different invariants relating the state of each process and the values that
might have been chosen in the previous rounds, thus a given round must know
which past rounds are fast or normal rounds in order to take over the execution
safely. When adding more kinds of rounds such an approach will quickly become
unmanageable, as each type of rounds must have one case for each possible other
kind of round, making $O(n^2)$ cases to consider in total.

Our notion of Composable Generalized Consensus defines a common interface and a
common correctness property for a round, abstracting over the particular round
implementations and making all types of rounds compatible with each other.

Moreover, our interface for rounds enables a crucial optimization seen in
practical systems: introducing an external reconfiguration master which manages
the addition of new replicas and the removal of failed ones. Interestingly, as
explained in \cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication},
in a real system it is often possible to implement a reliable reconfiguration
master which allows to tolerate $f$ faults with $f+1$ replicas instead of the
traditional $2f+1$ replicas. The reconfiguration master can also make the choice
of the new type of round to execute, providing the system's reconfiguration
policy.
